**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-01-29 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract the analysis of the error profile

mock_error <- read.table(file="data/process/mock.error.report")
mock_error_nochim <- mock_error[mock_error$numparents==1,]
mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
```

@@

```{r error_profile}
mutations <- c("AA", "AT", "AG", "AC", "TA", "TT", "TG", "TC", "GA", "GT", "GG", "GC", "CA", "CT", "CG", "CC", "Ai", "Ti", "Gi", "Ci", "dA", "dT", "dG", "dC")

mutation_table <- mock_error_good[, which(colnames(mock_error_good)%in%mutations)]
mutation_count <- apply(mutation_table, 2, sum)

mutation_total <- sum(mutation_count)
matches <- sum(mutation_count[c("AA", "TT", "GG", "CC")])
mis_matches <- mutation_total - matches
init_error_rate <- 100 * mis_matches / mutation_total		#0.6514488

#substitutions
substitutions <- mutations[grep("[ATGC].", mutations)]
substitutions <- substitutions[grep(".[ATGC]", substitutions)]
substitutions <- substitutions[! substitutions %in% c("AA", "TT", "GG", "CC")]
substitution_counts <- mutation_count[substitutions]
substitution_total <- sum(substitution_counts)
substitution_bias <- substitution_counts / substitution_total

reference <- gsub(".(.)", "\\1", names(substitution_bias))
# aggregate(substitution_bias, by=list(reference), mean)
#
# Group.1          x
# 1       A 0.10192188
# 2       C 0.05944286
# 3       G 0.07434043
# 4       T 0.09762816

substitution_rate <- 100 * substitution_total / mis_matches

# [1] 50.90346

#deletions
deletion_counts <- mutation_count[grep("d.", mutations)]
deletion_total <- sum(deletion_counts)
deletion_bias <- 100 * deletion_counts / deletion_total

#dA        dT        dG        dC
#24.25770 17.99132 39.44415 18.30683

deletion_rate <- 100 * deletion_total / mis_matches

#[1] 17.93361

#insertions
insertion_counts <- mutation_count[grep(".i", mutations)]
insertion_total <- sum(insertion_counts)
insertion_bias <- 100 * insertion_counts / insertion_total

#Ai        Ti        Gi        Ci
#23.13509 19.63951 30.05906 27.16633

insertion_rate <- 100 * insertion_total / mis_matches

#[1] 31.16293

# Need to connect error types with the quality scores
error_quality <- read.table(file="data/process/mock.quality.report", header=T, row.names=1)

#get quantiles
matches <- rep(as.numeric(rownames(error_quality)), error_quality$matches)
m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
m.94 <- 100 * sum(matches >= 94)/length(matches)
# [1] 80.52649

subs <- rep(as.numeric(rownames(error_quality)), error_quality$substitutions)
s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
s.94 <- 100 * sum(subs >= 94)/length(subs)
# [1] 79.96451

ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
i.94 <- 100 * sum(ins >= 94)/length(ins)
# [1] 80.35283

@@

basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736
```

***The PacBio error profile and a basic sequence curation procedure.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
`r paste(range(c(matches, subs, ins)), collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).

@@

error_by_coverage <- function(region){
	coverage_region <- mock_error_basic[mock_error_basic$region == region,]
	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)
	coverage_error <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"x"]
	names(coverage_error) <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"Group.1"]
	coverage_error
}

@@

	deep <- region_data$coverage >= 10
	coverage_data <- region_data[deep,]

>>>>

library(json)
```

@@

```{r error_profile}
error_profile <- fromJSON(scan("data/process/error_profile.json", quiet=T, what="character"))

@@

basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736

***The PacBio error profile and a basic sequence curation procedure.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`,
`r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines
(`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted
than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines
(`r error_profile$del_bias$dT, digits=2, nsmall=1)`%). The PacBio quality values varied between
`r paste(error_profile$qual_range, collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$qual_error_corr, digits=2, nsmall=2)`; Figure 1A).

@@

error_by_coverage <- function(region){
	coverage_region <- mock_error_basic[mock_error_basic$region == region,]
	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)

	coverage_error <- rep(NA, max(mock_error_basic$coverage))
	coverage_error[coverage_error_agg$Group.1] <- coverage_error_agg[,"x"]
	coverage_error
}

@@

	deep <- region_data$coverage >= cutoff
	coverage_data <- region_data[deep,]
